---
title: "frazer_gad_assessment"
author: "Rufus Frazer"
date: "November 19, 2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
#load libraries needed
library(readr) #readtable
library(dplyr) #data aggregation etc
library(tidyr) #spread
library(data.table) #aggregate
library(ggplot2) #visuals

```


## Request

For each line, the first column represents the time the measurement was made, the second
column represents the id of the patient submitting the assessment, the third column is the type
of assessment submitted, the fourth column is the date that the patient was created, and the
fifth column represents the final score of the assessment.

Given the information you have and any light research you'd like to do on the topic, what
insights can you draw? What assumptions have you made about the data? What are 2-3
additional pieces of information that would be important to collect? We are not looking for
production-ready code, but we will assess both your approach to visualization and your
technical abilities.


## Contents

- Summary Descriptive Information
- Sample Visual

start: 11/19: 9:42AM
finish code part: 11/19: 11:28AM
sql effort: 25 min



```{r load_data, message=FALSE, warning=FALSE, include=FALSE}
#read in data
phq <- read_csv("C:/Users/Rufus/Downloads/phq_all_final.csv")

```

```{r cleanup_data, message=FALSE, warning=FALSE, include=FALSE}
#clenup and quick data understanding-- assume that data is relatively clean
# table(phq$type) #gad7 only #repeated for similar
# nlevels(as.factor(phq$patient_id)) #15502
#abandon the time stamp for date, for simplicity, assuming no issues w this in data
phq$date<-as.Date(phq$date)
phq$patient_date_created<-as.Date(phq$patient_date_created)

# table(phq$first_form) #find out how many tests are first tests
#how many tests were taken by the same patient on the same day
test_cts<-as.data.table(phq)[,.(ct = .N), by = c("patient_id","date")]
table(test_cts$ct) #tests taken by single patients on the same date - shows multipel onthe same day

phq<-phq %>% 
        group_by(patient_id,date,patient_date_created) %>%  #don't need tyupe, all gad
        summarize(score = mean(score)) %>% #average by date, given that some may take multiple tests on the same day
      ungroup()
```

```{r add_fields, message=FALSE, warning=FALSE, include=FALSE}
#add some summary columns

#set a column for patient test number
phq<-phq %>%
      group_by(patient_id) %>% 
      arrange(date) %>%
      mutate(test_num = rank(date)) %>% #add a test number
      ungroup()

#set a column for level of gad, based on instructions description:
# 0-5 low 
# 6-10 mild
# 11-15 moderate
# 16 + severe
phq$severity_level_bin<-0 #init-- should be nonne remaining-- catch data issues
phq$severity_level_bin[phq$score<= 5]<-'low' #1 #
phq$severity_level_bin[phq$score > 5 & phq$score<= 10]<- 'mild'#2 #
phq$severity_level_bin[phq$score > 10 & phq$score<= 155]<- 'moderate'#3 # 
phq$severity_level_bin[phq$score > 15]<-'severe' #4 #
table(phq$severity_level_bin)
phq$possible_disorder_bin<-ifelse(phq$score >= 10,"disorder","no disorder") #(1,0)

#create a table w whether or not the data has a patients first form
phq$first_form<-ifelse(phq$date==phq$patient_date_created, 1,0) #see if test was done on same day as create day
first_form<-phq %>% group_by(patient_id) %>% summarize(first_form = max(first_form))

```


```{r patient_summary, message=FALSE, warning=FALSE, include=FALSE}
#some quick insights

#count of patients w first visits in dataset
first_vis<-table(first_form$first_form)
#count of patients with potential gad-- patient level
potential_gad<-table(phq$possible_disorder_bin)

#set a value of relevant change-- where it counts as "improving"
chg <-3

#count of patients by worst scenario-- patient level
potential_gad_cts<-phq %>% 
               group_by(patient_id) %>% 
               summarize(max_severity_level = max(severity_level_bin),
                         max_possible_disorder = max(possible_disorder_bin)) %>%
                ungroup() 
patients_by_severity<-table(potential_gad_cts$max_severity_level) #count of patients by worst reported severity
patients_by_gad<-table(potential_gad_cts$max_possible_disorder) #count of patients by worst reported severity

#count of patients by overall improvement-- earliest available visit better/worse than last available visit
improving_patients<-phq %>% 
               group_by(patient_id) %>% 
               filter(test_num == max(test_num) | test_num == 1.0)  %>% #first or last visit
               ungroup() %>%
                mutate(visit_rank = ifelse(test_num == 1.0,'first','last')) %>% #set whether its a first or last visit
               select(patient_id,score,visit_rank) %>% #choose columns
              spread(key = visit_rank, value = score) %>%
              filter(!is.na(last)) %>% #skip those who have only had one test
              mutate(improving_patient = ifelse(first > last + chg,'improving', #improving
                                                ifelse(first < last - chg,"declining", #declining
                                                       "no change"))) #same
improving_patients<-merge(improving_patients,first_form, by = "patient_id")
#et summary on patient progress
#imprpoving by wehther data has first form in it
table(improving_patients$improving_patient,improving_patients$first_form)
#improving by starting severity level
patient_severity_start<-phq %>% filter(test_num == 1.0) %>% 
  mutate(severity_level_start = severity_level_bin) %>%
  select(patient_id,severity_level_start)
improving_patients<-merge(improving_patients,patient_severity_start, by = "patient_id")

table(improving_patients$improving_patient,improving_patients$severity_level_start)



```

## Answering questions


Q:Given the information you have and any light research you'd like to do on the topic, what insights can you draw? 

Likely can make the argument that this group has improved over time

Q:What assumptions have you made about the data?
Lots: that it's clean-- no obvious errors of user input or data quality; that it's complete-- that patients have all tests included for a given time period andd that time period is a relevan timme period. I might also guess that some treatment has happened, but it would be good to know if this is a control or treatment sample- have these patients had any intervention?


Q:What are 2-3additional pieces of information that would be important to collect? 
Baseline expectations (both about the survey audience-- is this a population sample-- and treatment if any) and typical outcomes could help set up the evaluation of these data. Where does this data come from? Is the patient group homogenous? 




```{r patient_visuals, message=FALSE, warning=FALSE, include=FALSE}
#visuals

#do some basic visuals on the above items
hl<-as.data.table(improving_patients)[,.(ct = uniqueN(patient_id)), by = c("improving_patient","severity_level_start")]
hl<-hl %>% group_by(severity_level_start) %>% mutate(pct = ct/sum(ct)) %>% ungroup()
hl$pct<-round(hl$pct,3)


```


## Patient outcomes are positive for severe cases
```{r display_visuals, echo=FALSE}
health_improvement<- ggplot(data = hl, aes(x = severity_level_start, y = pct, fill = improving_patient)) +     geom_bar(stat = "identity",color = "black", position = position_dodge()) + 
   scale_y_continuous(labels=scales::percent) +
  theme_minimal() +
# theme(text = element_text(size=12)) +
        labs(title = 
              "Improvement Rates in Patients "
                  ,
         subtitle =   "by Starting Several Level"
          ,
         caption = "Improvement or Decline based on a change of 3 or more points on the GAD7 test"
         ,
         y = "Patients", x = "Starting Severity")

health_improvement

```


## SQL answers

Note; these are written blind, so it's possible that syntax errors exist. they assume clean tables, eg no repeating values etc

## query 1
1. how many users compplete an exercise in their first month per monthly cohort? (users: user_id,created_at; exercises: exercise_id,user-id,exercise_completion_date

with cohortss as
--get user_id and start month
select user_id,min(concat(extract ( year from created_at), extract(month from created_at))) cohort from users
group by user_id
,--get user_id exercise completion month
completion as
select  
 user_id,concat(extract ( year from created_at), extract(month from created_at) yrmon, 1 as exercises
group by 1,2

select user.cohort, sum(exercises.exercises)/count(*)  --number who completed/number who started
from 
cohort
left join completion
on cohort.user-id = completion.user_id
and user.cohort = completion.yrmon
group by user.cohort
## query 2

2. frequency dist for number of activities each user completed

--number each user completed
with exercises as(
select user_id, count(distinct exercise_id) exercises
from exercises
group by user_id)
select quartile, count(*) ct from (
select 
user_id,ntile(4) over(order by exercises) quartile
from exercises)
## query 3
3. Which orgs have most severe patient pop?
providers: provider_id,organization_id,organization_name
phq9: patient_id, provider_id, score, datetime_created
find top 5 orgs by avg score 

select org_name, avg(score) score from (
select ph.patient_id, org.org_name, score from providers org inner join phq9 ph
on ph.provider_id = org.provider_id)
group by org_name order by avg(score) desc limit 5
